---
title: "misc"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{misc}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Start by loading the normtest package

```{r setup}
library(normtest)
```

This vignette is to demonstrate some of the miscellaneous functionality of the normtest package.

## Example 1: You want to sort your data according to it's distribution and scedasticity, but you don't want to use the statistical tests provided here. 

The test_norm and two_group_test_norm functions are available to use independently of the 
statistical tests. Here's how to use them. 

You will still need a sample_info object for your data. You can see more details about this 
in the vignettes "multi_group_vignette" and "two_group_vignette". 

In the case of multiple groups (more than 2):

You'll have some data: 

```{r eval = TRUE}
mg_counts <- read.csv("multigroup_counts.csv")
#change first column to rownames
mg_counts <- cf(mg_counts)
head(mg_counts)
```

And you'll create your sample_info object:

```{r eval = TRUE}
mg_sample_info <- data.frame(samples = colnames(mg_counts), 
                          diet = c(rep("d1", 10), rep("d2", 10)), 
                          treatment = c(rep("t1", 5), rep("t2", 5), rep("t1", 5), rep("t2", 5)),
                          group = c(rep("d1t1", 5), rep("d1t2", 5), rep("d2t1", 5), rep("d2t2", 5)))
mg_sample_info
```

Then you can call test_norm:

```{r eval = FALSE}
res <- test_norm(mg_counts, mg_sample_info$diet, mg_sample_info$treatment)
```

Then you can access the three dataframes by: 

```{r}
#normal equal variance
ne <- res[[1]]
head(ne)
#normal nonequal variance
nu <- res[[2]]
head(nu)
#nonnormal
nn <- res[[3]]
head(nn)
```

In the case of two groups: 

You will again have some data:

```{r eval = TRUE}
tg_counts <- read.csv("twogroup_counts.csv")
#change first column to rownames
tg_counts <- cf(tg_counts)
head(tg_counts)
```

And your sample_info table:

```{r eval = TRUE}
tg_sample_info <- data.frame(sample = colnames(tg_counts), group = c(rep("GroupX", 4), rep("GroupY", 4)))
tg_sample_info
```

Then you can call two_group_test_norm:

```{r eval = FALSE}
res <- two_group_test_norm(tg_counts, tg_sample_info$group)
```

And you can access the individual dataframe exactly the same as above.

## Example 2: You already know your data to be suitable for a _____ test and you only want to do that test and maybe the multi comparison test.

The statistical tests are also available to use independently, and the process of using
them is largely the same. 

The most different of the tests is the two way anova test, which is only suitable for a 
multi group case. This is because it uses two experimental factors rather than one. 

We can use it like this: 

```{r eval = FALSE}
anova_results <- run_anova(mg_counts, mg_sample_info$diet, mg_sample_info$treatment)
```

If you wanted to do the Tukey multicomparison test, you can do it like this:

```{r eval = FALSE}
tukey_results <- run_tukey(mg_counts, mg_sample_info$group)
```

The rest of the tests are used just like the Tukey test. 

```{r eval = FALSE}
ttest_results <- run_ttest(tg_counts, tg_sample_info$group)
welch_results <- run_welch(mg_counts, mg_sample_info$group)
#the same run_welch function is used for both two group and multi group cases
kw_results <- run_kruskal(mg_counts, mg_sample_info$group)
```

etc...

## Example 3: You only want to calculate fold changes.

The same fold_change function is used for both multi and two group cases. 

```{r eval = TRUE}
mg_fc <- fold_change(head(mg_counts), mg_sample_info$group)
mg_fc
```

The two group case works the same way.



