---
title: "multi_group_vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{multi_group_vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Start by loading the normtest package

```{r setup}
library(normtest)
```

# Multi Group Workflow
What follows in this vignette is a standard workflow for using the normtest package to analyze data from
more than two groups. 
While it is possible to input raw data, it is recommended to use preprocessed, normalized, and cleaned
data. 

Our goal here is to take a dataset from an experiment with multiple factors and analyze it using the appropriate
statistics. 

We start by loading our data. We have a .csv file of normalized gene counts, called multigroup_counts.csv.

```{r loadData, eval = TRUE}
counts <- read.csv("multigroup_counts.csv")
head(counts)
```

Note that the data is already cleaned and ready for analysis. We do, however, need to change the column "X" 
to be the rownames of our data, and then remove the column. We use the cf() function included in the normtest
package. This is a very simple helper function that will set a specified column (defaults to column 1) to 
be rownames, and then remove the column from the data. 

```{r cf, eval = TRUE}
counts <- cf(counts)
head(counts)
```

Now we must build a sample_info object. The purpose of this object is to define which groups or other factors
each sample in our data belongs to. 

```{r sampleInfo, eval = TRUE}
sample_info <- data.frame(samples = colnames(counts), 
                          diet = c(rep("d1", 10), rep("d2", 10)), 
                          treatment = c(rep("t1", 5), rep("t2", 5), rep("t1", 5), rep("t2", 5)),
                          group = c(rep("d1t1", 5), rep("d1t2", 5), rep("d2t1", 5), rep("d2t2", 5)))
sample_info
```

Each sample from this experiement was given one of two diets, d1 or d2. Each sample was also given a treatment,
either t1 or t2. We have therefore constructed groups, which define both the diet and treatment of the sample.

The groups variable must be a linear combination of the other two variables (in this case, diet and treatment).
In other words, the group variable must describe the distribution of the other two variables.
group-d1t1 describes samples that got diet1 and treatment1, group-d2t1 describes samples that got diet2 and treatment1, etc.
The reason for this is that the multicomparison tests used in the analysis are not able to consider two 
factors, but rather only one. The statistical criteria used to sort the data into one of the three sets 
relies on the group distributions as defined by the first two variables. In order for these criteria to 
remain satisfied when performing the multicomparison tests, the group variable must follow this convention. 

The sample_info table above is an example of this. Below is a non-example (i.e., don't do this).

```{r notSampleInfo, eval = TRUE}
#dont do this!!! this is bad!!!
bad_sample_info <- data.frame(samples = colnames(counts), 
                          diet = c(rep("d1", 10), rep("d2", 10)), 
                          treatment = c(rep("t1", 5), rep("t2", 5), rep("t1", 5), rep("t2", 5)),
                          group = c(rep("group1", 2), rep("group2", 8), rep("group3", 3), rep("group4", 7)))
#dont use this!!! 
bad_sample_info
```

Notice how to group variables do not "line up with" or "describe" the distribution of the diet and 
treatment variables.

Now that we have our sample_info object and our counts ready, we can run the analysis. 
We can use the main function "rui" to call all the appropriate functions in succession. 

```{r callMain, eval = FALSE}
rui(counts, sample_info$diet, sample_info$treatment, sample_info$group)
```

This will write three .csv files in the current directory, "anova_results.csv", "multi_group_welch_results.csv", 
and "KW_results.csv". 

Also note that depending on the size of your data, this function could take several minutes to run. 

From here, the rest lies in the hands of the reader, but we will demonstrate what next steps are useful for our purposes.

Let's look at our anova_results.
```{r anovaAnalysis, eval = TRUE}
x <- read.csv("anova_results.csv")
x <- cf(x)
head(x)
```

This file contains data from every comparison, and it is unwieldy to consider every
comparison at once.

For this reason normtest includes the chop() function. This function can take one of the multi group output files, split it up into 
separate comparisons, and identify significant results by pvalue and fold change. To use it, we first must remove the columns not 
pertaining to a specific comparison from x. This is the first three columns.

```{r xTrim, eval = TRUE}
x1 <- x[, -c(1:3)]
head(x1)
```
Now x only contains pvalues and fold changes for each head-to-head comparision. Now we apply chop, the default pvalue cut off is 0.05, and 
the default absolute value fold change is 1.

```{r chop, eval = TRUE}
x_chopped <- chop(x1)
```

Note that x_chopped is now a list with two elements, both of which are also lists. The first element of x_chopped is a list of all the 
individual comparisons, which is our case is six. The second element of x_chopped is a list of only significant genes from each of the six 
comparisons. These are in the same order as the columns of x.

```{r listLengths}
length(x_chopped)
anova_res <- x_chopped[[1]]
length(anova_res)
anova_sigs <- x_chopped[[2]]
length(anova_sigs)
```

Suppose we're interested in the d1t2_vs_d2t2 comparison. Since this is the last comparison in x, it will also be the last comparison in res 
and sig. 

```{r getComp}
#this is the results for every gene
anova_comp <- anova_res[[6]]
head(anova_comp)
#this is the results only for genes we consider significant
anova_comp_sig <- anova_sigs[[6]]
nrow(anova_comp_sig)
head(anova_comp_sig)
```

Lets consider another use case. We'll use the KW_results file this time.

```{r eval = TRUE}
kw <- read.csv("KW_results.csv")
kw <- cf(kw)
```
 
Suppose we are interested in looking at the comparison d1t1_vs_d2t2. We'll use chop to get it, but we also could just do: 

```{r eval = FALSE}
kw_comp <- kw[, 6:7]
```

```{r KWchop}
#first remove the column that doesnt pertain to a specific comparison, this is only the first column for kw_results.
kw1 <- kw[, -1]
kw_chop <- chop(kw1)
kw_res <- kw_chop[[1]]
kw_sig <- kw_chop[[2]]
#and the d1t1_vs_d2t2 is third comparison
kw_comp <- kw_res[[3]]
head(kw_comp)
```
Then, maybe we want to see which fold changes are the largest (considering absolute values). 

```{r eval = TRUE}
ordered_kw <- kw_comp[order(abs(kw_comp[, 2]), decreasing = TRUE), ]
head(ordered_kw)
```

Certainly, negative infinity sounds like a large fold change. But this is actually likely
the result of us taking log2(0). So its probably more descriptive if we remove infinite 
values from this comparison. 

```{r eval = TRUE}
ordered_kw <- kw_comp[is.finite(kw_comp[, 2]), ]
ordered_kw <- ordered_kw[order(abs(ordered_kw[, 2]), decreasing = TRUE), ]
head(ordered_kw)
```

Now these values are more useful to us. Be aware that here we only looked at a subset of the d1t1_vs_d2t2 comparision;
the part that was suited to the Kruskal-Wallis test. To be thorough in our analysis we should look 
at the d1t1_vs_d2t2 comparison across all three results files.

We'll do that now, and then we'll use the results to create a heatmap of the 60 genes with largest fold changes and pvalues < 0.05. 
We'll use the pheatmap library.

```{r}
library(pheatmap)
```

```{r eval = TRUE}
#we already have the kw comparison, so now we need the correct anova comparison and welch
anova_comp <- anova_sigs[[3]]
we <- read.csv("multi_group_welch_results.csv")
we <- cf(we)
#apply chop
we1 <- we[, -1]
we_chopped <- chop(we1)
welch_sig <- we_chopped[[2]]
welch_comp <- welch_sig[[3]]
head(welch_comp)
```

Now we'll put the comparisons from the three different tests together into one dataframe, and sort it by fold change.

```{r eval = TRUE}
total <- rbind(anova_comp, welch_comp, kw_comp)
total <- total[order(total[, 2], decreasing = TRUE), ]
head(total)
```

And now we can make our heatmap. We want to plot the count data for these genes which we found to 
be significant at alpha = 0.05 and which have the largest fold changes. 

```{r}
#get the count data for these genes
total_names <- rownames(total)
total_counts <- counts[total_names, ]
head(total_counts)
```

```{r fig.height=6, fig.width=6}
#take log10 of each data point, adding 1 to account for any zeros
dat <- log10(total_counts + 1)
#we use part of our sample_info object to annotate the heatmap
anno <- cf(sample_info[, c(1,4)])
#make heatmap using top 60 genes by fold change
pheatmap(head(dat, 60), scale = "row", cluster_cols = FALSE, annotation_col = anno, fontsize = 6)
```
